Two subtle problems with over-representation analysis

### 2024-06-10 manuscript/manuscript.Rmd
ms

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
Despite this performance difference, a survey of articles published in 2019 showed that
seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
The discrepancy between performance and popularity is likely due to the relative ease of
conducting ORA as compared to FCS.
A minimal ORA might involve pasting a single list of genes into a text box on a website,
where the results appear nearly instantaneously.
On the other hand, FCS typically involves installing specific softwares and dealing with a
dataset representing every detected gene (~20,000 rows) and ensuring that the file
format from upstream tools is compatible with FCS packages.

Despite the popularity of ORA, there are concerns that this technique is being misused.
Due to technological and biological reasons, not all genes are detected in transcriptomic
studies, meaning that some genes will more readily appear as differentially expressed.
To address this bias, a background list of detected genes is required for comparison to the
list of selected genes (foreground list) [@Tilford2009-vx].
Failure to use a background gene list leads to dramatic changes in enrichment results,
rendering them unreliable [@Timmons2015-ki].
Unfortunately, use of an appropriate background list is reported in only a small fraction
(~4%) of peer-reviewed articles describing enrichment analysis [@Wijesooriya2022-fw].
Correction of p-values for multiple testing is crucial in controlling the false positive rate
as enrichment analysis typically involves throusands of parallel tests.
But appropriate p-value correction is reported in only ~54% of studies [@Wijesooriya2022-fw].
These issues have resulted in the development of best practices for end users and
stronger reporting standards [@Zhao2023-mx].

@@

The second problem is that adjusting p-values for multiple testing, also known as
false-discovery rate correction is implemented improperly.
Determining which pathways should be subject to false discovery rate correction depends on
how a detection threshold is set.
Ideally, a pathway should be considered detected based on the presence of a predetermined
number of member genes in the background set.
However some tools use the foreground set to define a pathway as detected.
This is problematic, as the act of trying to find an intersect between a foreground gene list
and a pathway is in itself a test, even if there are no common genes.
This results in pathways being discarded from the analysis if no common genes are found in
the foreground, despite a large number of those genes being present in the background list.
This artificially reduces the number of tests conducted and makes FDR values appear lower
than they should, potentially raising the rate of false positives.
We expect this problem to have a more severe effect when the foreground list is fairly small.

@@

Human gene sets were obtained from MSigDB v2023.2 [@Liberzon2015-cs], and included
KEGG Medicus [@Kanehisa2023-ni],
Reactome [@Milacic2024-iv],
Wikipathways [@Agrawal2024-lr],
MicroRNA targets from miRdb [@Chen2020-wo],
transcription factor targets from GTRD [@Kolmykov2021-mn],
Gene Ontology (GO) [@The_Gene_Ontology_Consortium2023-xp],
Human Phenotype Ontology (HPO) [@Kohler2021-ha],
Cell markers and
Hallmark pathways [@Liberzon2015-cs].
For an initial demonstration of the background problem, differentially expressed genes with
FDR<0.05 were subset, with separate lists for up and downregulated genes.
If fewer than 200 genes met this criteria, then the 200 genes with the smallest p-values
were selected.
These gene lists were subjected to ORA using the `enricher` function of clusterProfiler
v4.10.0, using a minimum set size of 5 and a no maximum set size.

For one test, a GO gene set called "Complement activation classical pathway" was added
to the KEGG database to test whether fold enrichment and p-values were affected.

@@

For this test, 2000 of the top up and down regulated genes were selected from each
differential expression dataset.
A workaround for the FDR problem was devised by filtering the gene sets by the presence of
at least two genes in the background list prior to running `enricher`.
After `enricher` using the parameters above, the number of gene sets in the results was
quantified, and compared to the number that should have been reported using the two gene
limit.
The difference between these numbers (*n*) represents the number of missing gene sets.
To account for these tests, *n* values of 1 were appended to the p-values obtained by
enricher, followed by FDR correction in R, to obtain the properly corrected FDR values.
As above, analysis of seven datasets was done with nine gene set libraries, the significant
sets were selected and the Jaccard statistics were collected for each run to compare original
and corrected analysis.

>>>>

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].
Despite this performance difference, a survey of articles published in 2019 showed that
seven of the top eight most popular tools were based on ORA [@Wijesooriya2022-fw].
The discrepancy between performance and popularity is likely due to the relative ease of
conducting ORA as compared to FCS.
A minimal ORA might involve pasting a list of gene symbols into a text box on a website,
where the results appear nearly instantaneously.
On the other hand, FCS typically involves installing specific softwares and dealing with a
dataset representing every detected gene (~20,000 rows) and ensuring that the file
format from upstream tools is compatible with FCS packages.

Despite the popularity of ORA, there are concerns that this technique is being misused.
Due to technological and biological reasons, not all genes are detected in transcriptomic
studies, meaning that some genes will more readily appear as differentially expressed.
To address this bias, a background list of detected genes is required for comparison to the
list of selected genes (foreground list) [@Tilford2009-vx].
Failure to use a background gene list leads to dramatic changes in enrichment results,
rendering them unreliable [@Timmons2015-ki].
Unfortunately, use of an appropriate background list is reported in only a small fraction
(~4%) of peer-reviewed articles describing enrichment analysis [@Wijesooriya2022-fw].
Correction of p-values for multiple testing is also crucial in controlling the false positive rate
as enrichment analysis typically involves throusands of parallel tests.
But appropriate p-value correction is reported in only ~54% of studies [@Wijesooriya2022-fw].
These issues have resulted in the development of best practices for end users and
stronger reporting standards [@Zhao2023-mx].

@@

The second problem is that adjusting p-values for multiple testing, also known as
false-discovery rate correction is implemented improperly.
Determining which pathways should be subject to false discovery rate correction depends on
how a detection threshold is set.
Ideally, a pathway should be considered detected based on the presence of a predetermined
number of member genes in the background set.
However some tools use the foreground set to define a pathway as detected.
This is problematic, as the act of trying to find an intersect between a foreground gene list
and a pathway is in itself a test, even if there are no common genes.
This results in pathways being discarded from the analysis if no common genes are found in
the foreground, despite a large number of those genes being present in the background list.
This artificially reduces the number of tests conducted and makes FDR values appear lower
than they should, potentially raising the rate of false positives.
We expect this problem to have a more severe effect when the foreground list is small.

@@

Human gene sets were obtained from MSigDB v2023.2 [@Liberzon2015-cs], and included
KEGG Medicus [@Kanehisa2023-ni],
Reactome [@Milacic2024-iv],
Wikipathways [@Agrawal2024-lr],
MicroRNA targets from miRdb [@Chen2020-wo],
transcription factor targets from GTRD [@Kolmykov2021-mn],
Gene Ontology (GO) [@The_Gene_Ontology_Consortium2023-xp],
Human Phenotype Ontology (HPO) [@Kohler2021-ha],
Cell markers and
Hallmark pathways [@Liberzon2015-cs].
To demonstrate the background problem, differentially expressed genes with
FDR<0.05 were subset, with separate lists for up and downregulated genes.
If fewer than 200 genes met this criteria, then the 200 genes with the smallest p-values
were selected, as suggested by a previous report [@Tarca2013-xh].
These gene lists were subjected to ORA using the `enricher` function of clusterProfiler
v4.10.0, using a minimum set size of 5 and a no maximum set size.

@@

For this test, 2000 of the top up and down regulated genes were selected from each
differential expression dataset based on p-value.
A workaround for the FDR problem was devised by filtering the gene sets by the presence of
at least two genes in the background list prior to running `enricher`.
After `enricher` using the parameters above, the number of gene sets in the results was
quantified, and compared to the number that should have been reported using the two gene
limit.
The difference between these numbers (*n*) represents the number of missing gene sets.
To account for these tests, *n* values of 1 were appended to the p-values obtained by
enricher, followed by FDR correction in R, to obtain the properly corrected FDR values.
As above, analysis of seven datasets was done with nine gene set libraries, the significant
sets were selected and the Jaccard statistics were collected for each run to compare original
and corrected analysis.
